Starcode: sequence clustering based on all-pairs search

### 2014-11-06 doc/document.tex
Updated benchmark with TRiP data and results text.

The running time of the different tools as a function of the size
of the clusters is shown on Figure \ref{fig:benchmark}c. The running
time of slidesort and starcode show a linear and sub-linear increase,
respectively. Seed and cd-hit run approximately in constant time
regardless of the cluster size. In spite of this trend, the
performance of starcode remains competitive, even for clusters of 1
million sequences. The memory usage is shown in Figure
\ref{fig:benchmark}c. The smallest memory footprint is achieved by
slidesort and cd-hit, with a maximum difference of an order of
magnitude with respect to the other tools. Comparing with
slidesort is not completely fair since it does not hold in memory the
full graph necessary for clustering, which is by far the most memory
demanding structure. The memory usage of starcode is the highest for
clusters of size 1000, but it decreases and becomes lower than the
memory usage of seed as the size of the cluster increases.

@@

The basic properties of the datasets used for benchmarking are
summarized in Table~\ref{tab:biobench_properties}. Datasets 1 and 2
consist of Illumina single reads; they differ by the total read count
and by the empirical cluster sizes. According to the output of starcode,
the largest clusters of dataset 1 contain approximately 2,000 sequences,
while dataset 2 contains 4 clusters with more than 1 million sequences.
Dataset 3 consists of paired-end reads.

\begin{table}[h]
\centering
\caption{Summary of the biological datasets used for benchmarking.
All the datasets are Illumina reads.}
\begin{tabular}{l|l c r}
    \textit{dataset}   & \textit{read count} &
    \textit{read length} & \textit{type} \\ \hline
    dataset 1 & 16,457,527   & 50  & single      \\
    dataset 2 & 127,675,537  & 50  & single      \\
    dataset 3 & 2,460,226    & 50+50  & paired-end \\
\end{tabular}
\label{tab:biobench_properties}
\end{table}

The running times of starcode, seed, slidesort, rainbow and cd-hit
are summarized in Table~\ref{tab:biobench_runtime}. Starcode was faster
than the other tools on all the datasets. Seed came in second position
on datasets 1 and 2, with a running time approximately 800 and 20 times
greater, respectively. Rainbow came in second position on dataset 3,
but it is the only other tool able to cluster paired-end reads. We did not
record the exact running times past 10 days since this is several
orders of magnitude higher than the running time of starcode.

\begin{table}[h]
\centering
\caption{Running time (in seconds) of the software on three biological
datasets. Runs longer than 10 days were interrupted. A dash indicates
that the software cannot be used for this dataset.}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{dataset 1} &
      \textit{dataset 2} & \textit{dataset 3} \\ \hline
    starcode   & 171         & 2,898       & 44    \\
    seed       & 135,959     & 60,374      & -     \\
    sidesort   & 632,168     & $>$ 10 days & -     \\
    rainbow    & -           & -           & 306   \\
    cd-hit-est & $>$ 10 days & 512,591     & -     \\
\end{tabular}
\label{tab:biobench_runtime}
\end{table}

The memory footprint of the different tools on the same datasets is
shown in Table~\ref{tab:biobench_memory}. The values represent the
peak memory usage throughout the run on the datasets described above.
On datasets 1 and 3, starcode had a significantly larger memory usage
than the best tool, while on dataset 2 it used significantly less
memory than seed and approximately as much as cd-hit.

\begin{table}[h]
\centering
\caption{Memory usage (in GB).}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{dataset 1} &
      \textit{dataset 2} & \textit{dataset 3} \\ \hline
    starcode   & 9.4 & 30.9 & 5.2 \\
    seed       & 4.1 & 53.9 & -   \\
    sidesort   & 1.9 & 13.9 & -   \\
    rainbow    & -   & -  & 0.5   \\
    cd-hit-est & 3.8 & 28.5 & -   \\
\end{tabular}
\label{tab:biobench_memory}
\end{table}

>>>>

The running time of the different tools as a function of the size
of the clusters is shown on Figure \ref{fig:benchmark}c. The running
time of slidesort and starcode show a linear and sub-linear trends,
respectively. Seed and cd-hit run approximately in constant time
regardless of the cluster size. In spite of this result, the
performance of starcode remains competitive, even for clusters of 1
million sequences. The memory usage is shown in Figure
\ref{fig:benchmark}c. The smallest memory footprint is achieved by
slidesort and cd-hit, with a maximum difference of an order of
magnitude with respect to the other tools. Note that the comparison with
slidesort is not completely fair since it does not hold in memory the
full graph necessary for clustering (which is by far the most memory
demanding structure). The memory usage of starcode is the highest for
clusters of size 1000, but it decreases and becomes lower than the
memory usage of seed as the size of the cluster increases.

@@

The basic properties of the datasets used for benchmarking are
summarized in Table~\ref{tab:biobench_properties}. Dataset 1
(SRR950457) has been pre-processed to extract the barcode and remove
the constant part of the reads. Only barcodes between 15 and 17
nucleotides were included in the file. Dataset 2 (SRRHENCHI) consists
of raw Illumina single reads. These datasets differ by the read size,
the total read count and by the empirical cluster sizes. According to
the output of starcode, the largest clusters of dataset 1 contain
approximately 70,000 sequences, whereas dataset 2 contains 4 clusters
with more than 1 million sequences. Dataset 3 (SRR950477) has been
included to benchmark starcode against rainbow in paired-end
clustering mode.
\begin{table}[h]
\centering
\caption{Summary of the biological datasets used for benchmarking.
All the datasets are Illumina reads.}
\begin{tabular}{l|l c r}
    \textit{dataset}   & \textit{read count} &
    \textit{read length} & \textit{type} \\ \hline
    SRR950457 & 6,542,309    & 16     & single     \\
    SRRxxxxxx & 127,675,537  & 50     & single     \\
    SRR950477 & 2,460,226    & 50+50  & paired-end \\
\end{tabular}
\label{tab:biobench_properties}
\end{table}

The running times of starcode, seed, slidesort, rainbow and cd-hit
are summarized in Table~\ref{tab:biobench_runtime}. We accommodated the
distance threshold for the first dataset to compensate for the reduced
sequence length. Both starcode and slidesort were executed with
the option '-d 2' and the identity for cd-hit was set to '-c 0.85'. We
were not able to run seed on dataset 1 due to limitations on the
minimum input sequence length. Starcode was faster than the other
tools on all the datasets. CD-HIT and seed came in second position
with a running time approximately 35 and 20 times greater on datasets
1 and 2, respectively. Rainbow was nearly an order of magnitude slower
in the job of clustering paired-end reads. We did not record the exact
running times past 10 days since this is several orders of magnitude
higher than the running time of starcode.

\begin{table}[h]
\centering
\caption{Running time (in seconds) of the software on three biological
datasets. Runs longer than 10 days were interrupted. A dash indicates
that the software cannot be used for this dataset.}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{SRR950457} &
      \textit{SRRxxxxxx} & \textit{SRR950477} \\ \hline
    starcode   & 5           & 2,898       & 44    \\
    seed       & -           & 60,374      & -     \\
    sidesort   & 4,055       & $>$ 10 days & -     \\
    rainbow    & -           & -           & 306   \\
    cd-hit-est & 170         & 512,591     & -     \\
\end{tabular}
\label{tab:biobench_runtime}
\end{table}

The memory footprint of the different tools on the same datasets is
shown in Table~\ref{tab:biobench_memory}. The values represent the
peak memory usage throughout the run on the datasets described
above. On short reads (dataset 1), starcode outperforms the other
tools taking advantage of the trie compaction. On dataset 3, starcode
had a significantly larger memory usage than rainbow. Starcode and
cd-hit used the same amount of memory on dataset 2. Both needed twice
as much memory as slidesort, which runs in advantage because it does
not need to keep the complete graph during the all-pair comparison
step to perform future clustering.

\begin{table}[h]
\centering
\caption{Memory usage (in GB).}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{dataset 1} &
      \textit{dataset 2} & \textit{dataset 3} \\ \hline
    starcode   & 0.65 & 30.9 & 5.2 \\
    seed       & -    & 53.9 & -   \\
    sidesort   & 1.30 & 13.9 & -   \\
    rainbow    & -    & -    & 0.5 \\
    cd-hit-est & 0.80 & 28.5 & -   \\
\end{tabular}
\label{tab:biobench_memory}
\end{table}
